Two applications of logic programming to Coq

### 2020-11-03 papers/types2020/conc.tex
first pass to sec 3

\todo{sum up}

\begin{metanote}
  Decided against a related work section: most is already discussed in the intro, the rest can be mentioned in passing --am
\end{metanote}
% \subsection{Related work}
% \label{ssec:rel}

@@

\subsection{Future work}
\label{ssec:fut}

The code reported in Fig.~\ref{fig:augmented} is somewhat idealized in
so far that we omitted a rule performing weak-head reduction (allowing
to handle directly existential goals) and more importantly that used
the \emph{delay} mechanism. This a distinguished feature of Elpi, whose
usefulness has been advocated in ~\cite{DunchevCT16,coen19mscs}.  The
use of constraints for data generation is well
developed~\cite{FioravantiPS15} and we could try to leverage to
improve PBT using partially instantiated terms: for example generating
say lists with at least two elements as \texttt{[1,2|Tail]}, without
recurring to needed narrowing as in in
LazySmallCheck~\cite{smallcheck}.

\todo{Link to the repo? \url{https://github.com/proofcert/fpc-elpi}}

@@

Extending the LJF kernel to handle defined atoms
%%% Local Variables:
%%% mode: latex
%%% TeX-master: "main"
%%% End:

>>>>

\todo{sum up contribution}

% \begin{metanote}
%   Decided against a related work section: most is already discussed in the intro, the rest can be mentioned in passing --am
% \end{metanote}
% \subsection{Related work}
% \label{ssec:rel}

@@

% \subsection{Future work}
% \label{ssec:fut}
\todo{finish future work}
There are many avenues of developments for thisline of research.
The code reported in Fig.~\ref{fig:augmented} is somewhat idealized in
so far that we omitted a rule performing weak-head reduction (allowing
to handle directly existential goals) and more importantly that used
the \emph{delay} mechanism. This a distinguished feature of Elpi, whose
usefulness has been advocated in ~\cite{DunchevCT16,coen19mscs}.  The
use of constraints for data generation is well
developed~\cite{FioravantiPS15} and we could try to leverage to
improve PBT using partially instantiated terms: for example generating
say lists with at least two elements as \texttt{[1,2|Tail]}, without
recurring to needed narrowing as in in
LazySmallCheck~\cite{smallcheck}.

@@

Extending the LJF kernel to handle defined atoms

\todo{Link to the repo. Which one? \url{https://github.com/proofcert/fpc-elpi} is provate and full of stuff}
%%% Local Variables:
%%% mode: latex
%%% TeX-master: "main"
%%% End:

### 2020-11-03 papers/types2020/elab.tex
first pass to sec 3

If one is interested only in building Coq kernel proof structures, we
can bypass the use of an explicit pairing operation and build the
synthesis of such proof structures directly into the FPC proof
checker.  If one is interested in checking only one kind of external
proof structure, then the FPC for that structure could also be built
into the checker (via, say, partial evaluation of logic programs
\cite{lloyd91jlp}).

>>>>

\todo{Mention that we did \textbf{not} use pairing as well for the check predicate in sec 3}
If one is interested only in building Coq kernel proof structures, we
can bypass the use of an explicit pairing operation and build the
synthesis of such proof structures directly into the FPC proof
checker.  If one is interested in checking only one kind of external
proof structure, then the FPC for that structure could also be built
into the checker (via, say, partial evaluation of logic programs
\cite{lloyd91jlp}).

### 2020-11-03 papers/types2020/sec3.tex
first pass to sec 3

%\newpage
\section{Proof theory and proof certificates}

% {\color{red} The goal in this section is to say a bit about focusing
%   generally, and to give a bit of detail about FPC for Horn clauses.
%   Initial draft of text taken from \cite{blanco19ppdp}.
% }

\begin{metanote}
  (AM): Start with the logical reconstruction of \texttt{check}, to and then reverse engineer what we want/need to say about focusing and FPC. We \textbf{do} not talk about PBT yet, just logic programming (with proof terms). % In sec 4 we package \texttt{check} in a PBT query and explain how \texttt{check} works for term generation.
% One way to proceed:
  % \begin{itemize}
  % \item One figure fig \ref{fig:interp} for Horn logic over Coq terms (no certificates) and its encoding with quotations 
  % \item Then the logic behind the dependent kernel with certificates and proof terms
  % \end{itemize}
%
Stuff in red is from old paper
\end{metanote}

\subsection{Focused proof systems with certificates}
\label{ssec:focused}
\todo{Adapt the discussion}
\todo{grammar for formulae and proof terms \cite{Herbelin94,LengrandDM06}  or just the figure? Mention spine calculus}

The proof search approach to encoding Horn clause computation
results in the structuring of proofs with repeated
switching between a \emph{goal-reduction} phase and a
\emph{backchaining} phase~\cite{miller91apal}.

%
{\color{red}
The notion of \emph{focused proof systems}~\cite{andreoli92jlc,liang09tcs}
generalizes this view of proof construction by identifying the
following two phases. % of proof construction.
\begin{enumerate}
\item The \emph{negative}%
% \footnote{The terminology of negative and positive phases is a bit
%   unfortunate: historically, these terms do not refer to positive or negative
%   subformula occurrences but rather to certain semantic models used in
%   the study of linear logic~\cite{girard87tcs}.}
%
  phase corresponds to  goal-reduction: in this phase,
  inference rules that involve \emph{don't-care-non\-de\-ter\-min\-ism} are
  applied.
%
  As a result, there is no need to consider backtracking over choices
  made in building this phase.

\item The \emph{positive} phase corresponding to  backchaining: in this phase, inference rules that involve
  \emph{don't-know-non\-de\-ter\-min\-ism} are applied: here,
  inference rules need to be supplied with information in order to
  ensure that a completed proof can be built.
%
  That information can be items such as which term is needed to
  instantiate a universally quantified formula and which disjunct of a
  disjunctive goal formula should be proved.
\end{enumerate}
%
Thus, when building a proof tree (in the sequent calculus) from the
conclusion to its leaves, the negative phase corresponds to a simple
computation that needs no external information, while the positive
phase may need such external information to be supplied.
%
In the literature, it is common to refer to a repository of such
external information as either an \emph{oracle} or a \emph{proof
certificate}.
}

@@

\todo{Figure now uses Dale kernel-v2 --- needs further abstraction and explanation, see appendix}
\begin{footnotesize}
\begin{figure}
    \newcommand{\XXi}{{\color{blue}{\Xi}}}

  \newcommand{\bc}[5]{#1;#2\vdash #3 : #4 \Downarrow #5}
\[
% \infer{\XXi\vdash t = t}
% {%\eqExpert{\XXi}
% }
% \qquad
% \infer{\XXi\vdash \true}
% {%\trueExpert{\XXi}
% }
% \qquad
% \infer{\XXi\vdash G_1\vee G_2}
% {\XXi\vdash G_i%\qquad \orExpert{\XXi}{\XXi'}{i}
% }
% \qquad
% \infer{\XXi\vdash \exists x. G}
% {\XXi\vdash G[t/x]%\qquad \someExpert{\XXi}{\XXi'}{t}
% }
% \qquad 
\infer{\XXi;\Gamma\vdash \lambda x:A.\,t : \forall x : A.\,  G}
{\XXi;\Gamma, x : A\vdash t : G%\quad \andExpert{\XXi}{\XXi_1}{\XXi_2}
}
\qquad
\infer{\XXi;\Gamma\vdash t : A}
      {\bc{\XXi}{\Gamma} t D A \quad D \in \Pscr\cup\Gamma}
\]
\vskip -18pt
\[
  \infer{\bc {\XXi}{\Gamma} {[]} A A}{\trueExpert{\XXi}}{}\qquad
%  \infer{\bc{\forall x.\, D} A}{\bc{D[t/x]} A}{}\qquad
  \infer{\bc{\XXi}{\Gamma}{(t\ l)} {\forall x:B.\, D} A}
  {\XXi_1;\Gamma\vdash t : A\qquad \bc{\XXi_2}{\Gamma}l {B[t/x]} A\qquad\prodExpert{\XXi}{\XXi_1}{\XXi_2}}{}
\]
%\lstinputlisting[linerange={interp-end}]{code/dep-kernel.mod}
\caption{dep kernel.}
\label{fig:augmented}
\end{figure}
\begin{figure}
%\lstinputlisting[linerange={sigs-end}]{code/dep-kernel.sig}
\lstinputlisting[linerange={check-end}]{code/dep-kernel-v2.mod}
\caption{Implementation of a  dependent kernel for Coq terms.}
\label{fig:kernel}
\end{figure}
\end{footnotesize}

Figure~\ref{fig:augmented} contains a simple proof system \dots  in which each inference rule is augmented with an
additional premise involving an \emph{expert predicate}, a
certificate $\Xi$, and possibly continuations of certificates ($\Xi'$,
$\Xi_1$, $\Xi_2$) with extracted information from certificates (in the
case of $\vee$ and $\exists$).
%
%
%
\todo{Explain relations with other term assignments. Note that we do not use explicit substitutions (we are cut free). Explain the choice of not having certificates represent proof terms. Explain that the unfold clause handles defined atoms and defined connectives (conjunction, disjunction, existential and also equality): this may be less efficient and proof-theoretically induce pointless phase changes, but it is arguably very elegant e simplifies proof terms that arand inherited by Coq }

Figure~\ref{fig:kernel} contains the Elpi implementation of the
inference rules in Figure~\ref{fig:augmented}: here the infix
turnstile $\vdash$ symbol is replaced by the \lsti{check} predicate.
%
% Notice that it is easy to show that no matter how the expert predicates
% are defined, if the goal \lsti{check Cert B} is provable in \lP then
% \lsti{B} must be a sound consequence of the program clauses stored in
% \dots
% the \lsti{prog} predicate (which provides a natural implementation of
% the premise $(A~\hbox{\tt :-}~G)\in\instan{\Pscr}$.
% AM: not there
% in Figure~\ref{fig:kernel}).

%\subsection{Certificates}

\todo{What happened to the unfold certificate???}
% \todo{Show the dprolog tactic? have an example? could  be
% \lsti{Goal  exists R, insert 2 [0;1] R.} and we would move the code for insert here}
{\color{red}}
The notion of proof certificates
 is taken from the general setting of \emph{foundational
  proof certificates}~\cite{chihani17jar}.
%
In our case here, an FPC is a collection of \lP clauses that
provide the remaining details not supplied in Figure~\ref{fig:kernel}:
that is, the exact set of constructors for the \lsti{cert} type as
well as the exact specification of the expert predicates listed in
that figure.
%
Figure~\ref{fig:resources} displays two such FPCs,
both of which can be used to describe proofs where we bound
the number of occurrences of unfoldings in a proof.
%
For example, the first FPC
provides the experts for treating certificates that are constructed
using the \lsti{qheight} constructor.
%
As is easy to verify, the query \mbox{\lsti{(check (qheight 5) B)}} is
provable in  the kernel using the clauses in Figures~\ref{fig:kernel}
and~\ref{fig:resources} if and only if the height of that proof is 5
or less.
%
Similarly, the second FPC uses the constructor \lsti{qsize} (with two
integers) and can be used to bound the total number of instances of
unfoldings in a proof.
%
%In particular, the query \mbox{\lsti{(check (qsize 5 H) B)}}
% \begin{lstlisting}
% sigma H\ (check (qsize 5 H) B)
% \end{lstlisting}
%is provable if and only if the total number of unfoldings of that
%proof is 5 or less.

\begin{figure}
\lstinputlisting[linerange={resources-end}]{code/fpcs.sig}
\lstinputlisting[linerange={resources-end}]{code/fpcs.mod}
\caption{Two FPCs that describe proofs that are limited in either
  height or in size.}
\label{fig:resources}
\end{figure}

Further, if we view a particular FPC as a means of restricting proofs, it is
possible to build an FPC that \emph{restricts} proofs satisfying two FPCs
simultaneously.
%
% In particular, Figure~\ref{fig:pairing} defines an FPC based on the
% (infix) constructor \lsti{<c>}, which \emph{pairs} two terms of type
% \lsti{cert}.
%
The pairing experts for the certificate \lsti{Cert1 <c> Cert2} simply
request that the corresponding experts also succeed for both
\lsti{Cert1} and \lsti{Cert2} 
%
Thus, the query \lsti{check ((qheight 4) <c> (qsize 10)) B}
will succeed if there is a proof of \lsti{B} that has a height less
than or equal to 4 while also being of size less than or equal to 10.

@@

\subsection{A prolog-like tactic}

>>>>

%\newpage
\section{Proof theory and proof certificates}
\label{sec:three}
  \newcommand{\XXi}{{\color{blue}{\Xi}}}

\todo{Anything here?}
% \begin{metanote}
%   (AM): Start with the logical reconstruction of \texttt{check}, to and then reverse engineer what we want/need to say about focusing and FPC. We \textbf{do} not talk about PBT yet, just logic programming (with proof terms). % In sec 4 we package \texttt{check} in a PBT query and explain how \texttt{check} works for term generation.
% \end{metanote}

\subsection{Focused proof systems with certificates}
\label{ssec:focused}
\todo{Change title? }
\todo{Use same names in Fig and 3: A for Atom, L for ListArgs etc.}

% The proof search approach to encoding Horn clause computation
% results in the structuring of proofs with repeated
% switching between a \emph{goal-reduction} phase and a
% \emph{backchaining} phase~\cite{miller91apal}.

%

@@

% \todo{Figure now uses Dale kernel-v2 --- needs further abstraction and explanation, see appendix}
\begin{footnotesize}
\begin{figure}[t]
  %
  \renewcommand{\Gamma}{}
    \newcommand{\bc}[5]{#1#2\vdash #3 : #4 \Downarrow #5}
\[
% \infer{\XXi\vdash t = t}
% {%\eqExpert{\XXi}
% }
% \qquad
% \infer{\XXi\vdash \true}
% {%\trueExpert{\XXi}
% }
% \qquad
% \infer{\XXi\vdash G_1\vee G_2}
% {\XXi\vdash G_i%\qquad \orExpert{\XXi}{\XXi'}{i}
% }
% \qquad
% \infer{\XXi\vdash \exists x. G}
% {\XXi\vdash G[t/x]%\qquad \someExpert{\XXi}{\XXi'}{t}
% }
% \qquad 
\infer{\XXi\Gamma\vdash A : s}
{(A,s) \in \mathcal{A}
}
% \qquad
% \infer{\XXi;\Gamma\vdash \lambda x:A.\,t : \forall x : A.\,  G}
% {\XXi;\Gamma, x : A\vdash t : G%\quad \andExpert{\XXi}{\XXi_1}{\XXi_2}
% }
\qquad
\infer{\XXi_1\Gamma\vdash k\ l : A}
      {\bc{\XXi_2}{\Gamma} l D A \quad k : D \in \Pscr_A\cup\Gamma\qquad \unfoldExpert{\XXi_1}{\XXi_2}} 
\]
\vskip -18pt
\[
  \infer{\bc {\XXi}{\Gamma} {[]} A A}{\trueExpert{\XXi}}{}\qquad
%  \infer{\bc{\forall x.\, D} A}{\bc{D[t/x]} A}{}\qquad
  \infer{\bc{\XXi}{\Gamma}{(t\ ::\ l)} {\forall x:B.\, D} A}
  {\XXi_1;\Gamma\vdash t : A\qquad \bc{\XXi_2}{\Gamma}l {B[t/x]} A\qquad\prodExpert{\XXi}{\XXi_1}{\XXi_2}}{}
\]
%\lstinputlisting[linerange={interp-end}]{code/dep-kernel.mod}
\caption{dep kernel.}
\label{fig:augmented}
\end{figure}
\begin{figure}
\lstinputlisting[linerange={sigs-end}]{code/dep-kernel.sig}
\lstinputlisting[linerange={check-end}]{code/dep-kernel-v2.mod}
\caption{Implementation of a  dependent kernel for Coq terms.}
\label{fig:kernel}
\end{figure}
\end{footnotesize}

Figure~\ref{fig:augmented} contains an implicitely focused proof system  for a small fragment of the Calculus of Inductive Constructions that corresponds to Horn logic programming over pure \lsti{Inductive} definitions.
%
It is a sequent calculus inspired by the calculus for
proof search in Pure Type Systems introduced in \cite{LengrandDM06}, based in
 turn on ideas stemming from focusing (in particular, the LJT calculus of
Herbelin \cite{Herbelin94}). Similarly to that calculus, we have a term
language that includes terms and lists of terms, and two distinct typing
judgements for the two categories. In our case, where proof search steps are
viewed as computation, the term judgement corresponds to a goal-reduction
step while the list judgement corresponds to a backchaining step. This coincides
with the idea behind the \emph{spine calculus}~\cite{Cervesato97tr}.

Since we are only dealing with types that correspond to Horn clauses, and
atomic types are inductively defined, we do not need a $\forall$ rule on the right, although the proof theory would gladly allow it. This means, at the term level, that we
don't need to have variables in our grammar of terms. Terms are always
applied to a (possibly empty) list of arguments, and as is usual in Pure Type
Systems we have the same grammar for terms and types:
%, as shown in figure~\ref{fig:terms}.
\todo{Let's think about that: we need A at least, but A is also a metavariable for atoms}
%\begin{figure}
\begin{align*}
t,u ::= & k\ l \;|\; \forall x : t.\, u \\
l ::= & [] \;|\; (t\ ::\ l)
\end{align*}
% \caption{Grammar of terms}
% \label{fig:terms}
% \end{figure}

\todo{This does not see quite right: D vs A, consider say even}
The proof system is parametrized by  external set of inductive definitions available in
the (implict) global context: we denote by $\Pscr_A$ the set of type declarations for the
constructors of type $A$, of the form $k : D$ where $D$ is of the form
$\forall x_1 : t_1.\, \dots \forall x_n : t_n.\, D\ u_1 \dots u_m$. We denote
by $\mathcal{A}$ the set of arity declaration for the inductive types.

%
%
%done Explain relations with other term assignments. Note that we do not use
% explicit substitutions (we are cut free)
% Explain the choice of not having certificates represent proof terms.
\todo{  Explain that the unfold clause handles
defined atoms and defined connectives (conjunction, disjunction, existential
and also equality): this may be less efficient and proof-theoretically induce
pointless phase changes, but it is arguably very elegant e simplifies proof
terms that arand inherited by Coq }

Figure~\ref{fig:kernel} contains the Elpi implementation of the
inference rules in Figure~\ref{fig:augmented}: $\XXi\vdash t \colon A$ corresponds to
\lsti{check $\XXi$ (go A T)} and $\XXi\vdash l \colon D \Downarrow A$ corresponds to
\lsti{check $\XXi$ (bc D A L)}.

% here the infix
% turnstile $\vdash$ symbol is replaced by the \lsti{check} predicate, and every rule
% is directly mapped to a clause in the code. The argument of the clause is
% either given by a \lsti{go} or a \lsti{bc} constructor, signaling whether the clause
% is a goal-reduction or a backchaining one.

Coq terms are accessed through the Coq-Elpi API, and their representation
in \lP takes advantage of the native \lP constructs such as lists and binders.
The following is a list of some of signatures of the notions that we use:
\begin{lstlisting}
  kind term   type.   % reification of Coq terms
  kind gref    type     % reification of refernces to global terms
  type global gref -> term.
  type indt   inductive -> gref.  % reification of inductive types
  type indc   constructor -> gref. % reification of their constructors 
  type app    list term -> term.  % reification of nary application
  type prod   name -> term -> (term -> term) -> term.
\end{lstlisting}

Note that \lsti{prod} encodes dependent products taking a name for pretty printing,  a term and a \lP function from
terms to terms, namely $\forall x : A, B$ is encoded by \lsti{prod ``x'' A (x\ B x)}; when, in the implementation of the product-left rule, we
apply $B$ to the variable \lsti{Tm},
we get a term that can be used to continue backchaining. This
application is obtained via meta-level substitution, in the style of HOAS. In this sense, our calculus uses \emph{implict} substitutions, rather than explicit ones as in LJT and PTSC's tradition, the more since proof search in our application is cut-free. 
% Syntax: one can think as if something of type \lsti{term -> term} were a
% \lsti{term} with a hole in it for another \lsti{term}.
\todo{Now safe-dest-app is hiiden inside \texttt{definition}. Put the rest of the code in the appendix?}
The unfolding rule makes use of the Coq-Elpi primitives
\lsti{coq.safe-dest-app} in order to obtain the head term of a (possibly
nested) application, and \lsti{coq.env.indt} in order to access the global
environment of inductive definitions and query for information about them.

\smallskip
In figure~\ref{fig:augmented}  each
inference rule is augmented with an additional premise involving an
\emph{expert predicate}, a certificate $\Xi$, and possibly continuations of
certificates ($\Xi'$, $\Xi_1$, $\Xi_2$) with extracted information from
These notions are taken from the general setting of \emph{foundational
  proof certificates}~\cite{chihani17jar}.
%
In our case here, an FPC is a collection of \lP clauses that
provide the remaining details not supplied in Figure~\ref{fig:kernel}:
that is, the exact set of constructors for the  type of certificates as
well as the specification of the expert predicates listed \emph{ibidem}.
%
The top of Figure~\ref{fig:resources} displays two such FPCs,
both of which can be used to describe proofs where we bound
the dimension of  a proof.
%
For example, the first FPC dictates that the query \mbox{\lsti{(check (qheight 5) B)}} is
provable in  the kernel using the clauses in Figures~\ref{fig:kernel}
and~\ref{fig:resources} if and only if the height of that proof is 5
or less.
%
Similarly, the second FPC can be used to bound the total number of instances of
unfoldings in a proof.
%
%In particular, the query \mbox{\lsti{(check (qsize 5 H) B)}}
% \begin{lstlisting}
% sigma H\ (check (qsize 5 H) B)
% \end{lstlisting}
%is provable if and only if the total number of unfoldings of that
%proof is 5 or less.

\begin{figure}[t]
\lstinputlisting[linerange={resources-end}]{code/fpcs.sig}
\lstinputlisting[linerange={resources-end}]{code/fpcs.mod}
\caption{Sample FPCs}
\label{fig:resources}
\end{figure}

\todo{We could move here from elab the paragraph ``It is also possible to pair two different proof certificates, defined
by two different FPC definitions,'' and/or reduce redundancy}
Further, if we view a particular FPC as a means of restricting proofs, it is
possible to build an FPC that \emph{restricts} proofs satisfying two FPCs
simultaneously.
%
% In particular, Figure~\ref{fig:pairing} defines an FPC based on the
% (infix) constructor \lsti{<c>}, which \emph{pairs} two terms of type
% \lsti{cert}.
%
The pairing experts for the certificate \lsti{Cert1 <c> Cert2} simply
request that the corresponding experts also succeed for both
\lsti{Cert1} and \lsti{Cert2} 
%
Thus, the query \lsti{check ((qheight 4) <c> (qsize 10)) B}
will succeed if there is a proof of \lsti{B} that has a height less
than or equal to 4 while also being of size less than or equal to 10.

@@

\begin{metanote}
  AM stops here
\end{metanote}
\subsection{A prolog-like tactic}
